---
title: "EEG Spindle Analysis"
author: "Kevin Liu"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    number_sections: true
    keep_tex: true
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom)
library(effsize)
```

\newpage

# Analysis Overview

Features exhibiting more than 10% missingness were excluded, and median imputation was applied to remaining missing values. Independent t-tests were performed across all EEG features, adjusting for multiple comparisons using the Bonferroni correction method. Significant group differences (corrected p < 0.05) were identified for 139 EEG spindle and slow oscillation features. Cohen’s d was calculated for each significant feature to quantify the effect size and directionality of differences. The top 10 features with the largest effects were visualized using topographic EEG scalp maps created in Python (MNE library) to demonstrate distinct regional spindle and slow oscillation characteristics between ASD and healthy control groups.

\newpage

# Data Preparation

## Load and Merge Metadata

```{r initial_cleanup, message=FALSE}
# Load subject metadata
subject_data <- read_csv("data/eeg_metadata.csv") %>% 
  mutate(SID = sub("-.*$", "", SID)) %>% 
  select(SID, sex, age_year)

# Load spindle feature data and merge with subject metadata
spindle_data <- read_csv("data/features_spindles.csv") %>% 
  mutate(SID = sub("^(.*?)_.*$", "\\1", SID),
         SID = sub("-.*$", "", SID)) %>% 
  inner_join(subject_data, by = "SID") %>% 
  mutate(group = ifelse(Group == "TD", "HC", "ASD"),
         sex = ifelse(sex == 1, "M", "F")) %>% 
  select(SID, group, sex, age_year, everything(), -Group) %>% 
  column_to_rownames("SID")

# Remove rows with entirely missing data
spindle_data_usable <- spindle_data %>% 
  select(-group, -sex, -age_year) %>% 
  filter(if_any(everything(), ~ !is.na(.)))

# Keep only matched subjects with complete demographic data
spindle_data_matched <- spindle_data %>% 
  rownames_to_column("SID") %>% 
  filter(SID %in% rownames(spindle_data_usable), !is.na(sex)) %>% 
  column_to_rownames("SID")
```

\newpage

# Handling Missing Data

```{r handle_missing_data}
# Identify features with >10% missing data
gt10p_missing <- spindle_data_matched %>%
  summarize(across(everything(), ~mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Percent_Missing") %>%
  filter(Percent_Missing > 10) %>% 
  pull(Feature)

# Remove features and impute remaining missing data with median
spindle_data_filt_imp <- spindle_data_matched %>%
  select(-all_of(gt10p_missing)) %>% 
  mutate(across(where(is.numeric), ~ replace_na(., median(., na.rm = TRUE))),
         group = factor(group, levels = c("HC", "ASD")),
         sex = factor(sex, levels = c("F", "M")))

# Export the cleaned and imputed dataset
spindle_data_filt_imp %>% 
  rownames_to_column("SID") %>%
  select(SID, group, everything(), -sex, -age_year) %>% 
  write.csv("data/spindle_data_filtered_imputed.csv", row.names = FALSE)

# Check distribution by group and sex
table(spindle_data_filt_imp$group, spindle_data_filt_imp$sex)
```

\newpage

# Statistical Analysis

## Group-wise Comparisons (t-tests)

```{r t_statistics}
# Perform independent t-tests by group and adjust for multiple comparisons
signif_features <- spindle_data_filt_imp %>%
  select(-sex) %>% 
  pivot_longer(cols = -group, names_to = "feature", values_to = "value") %>%
  group_by(feature) %>%
  nest() %>%
  mutate(
    test = map(data, ~ t.test(value ~ group, data = .x)),
    tidied = map(test, tidy)
  ) %>%
  unnest(tidied) %>%
  select(feature, estimate1, estimate2, statistic, p.value) %>%
  rename(mean_HC = estimate1, mean_ASD = estimate2, t_stat = statistic, p_value = p.value) %>%
  mutate(p_adj = p.adjust(p_value, method = "bonferroni")) %>%
  filter(p_adj < 0.05) %>%
  arrange(p_adj)

signif_features
```

\newpage

## Effect Size Calculation (Cohen's d)

```{r effect_sizes}
# Calculate Cohen's d for each significant feature
cohens_d_results <- spindle_data_filt_imp %>% 
  pivot_longer(-c(sex, group), names_to = "feature", values_to = "value") %>%
  group_by(feature) %>%
  summarise(cohens_d = cohen.d(value ~ group)$estimate)

# Merge Cohen's d with significant features
final_ranking <- signif_features %>%
  left_join(cohens_d_results, by = "feature") %>%
  mutate(category = case_when(
    str_detect(feature, "AMP")        ~ "Amplitude",
    str_detect(feature, "DENS")       ~ "Density",
    str_detect(feature, "DUR")        ~ "Duration",
    str_detect(feature, "CHIRP")      ~ "Chirp",
    str_detect(feature, "COUPL")      ~ "Coupling",
    str_detect(feature, "ISA_M")      ~ "ISA per Minute",
    str_detect(feature, "ISA_S")      ~ "ISA per Spindle",
    str_detect(feature, "ISA_T")      ~ "Total ISA",
    str_detect(feature, "NOSC")       ~ "Oscillations per Spindle",
    str_detect(feature, "DISPERSION") ~ "Dispersion Index",
    str_detect(feature, "FFT")        ~ "Spindle Frequency (FFT)",
    str_detect(feature, "SYMM2")      ~ "Folded-Symmetry Metric",
    str_detect(feature, "SYMM")       ~ "Symmetry Metric",
    TRUE                              ~ "Other"
  )) %>%
  arrange(p_adj, desc(abs(cohens_d)))

# Export top 10 significant features
final_ranking %>% 
  head(10) %>% 
  write.csv("data/top10_significant_results.csv", row.names = FALSE)

head(final_ranking, 10)
```

\newpage

# EEG Visualization (Python & MNE)

```{python mne}
import pandas as pd
import numpy as np
import mne
import matplotlib.pyplot as plt

# Load cleaned EEG data
df = pd.read_csv("data/spindle_data_filtered_imputed.csv")

# EEG channel information
channels = ['Fp1', 'Fp2', 'F3', 'F4', 'C3', 'C4', 'P3', 'P4',
            'O1', 'O2', 'F7', 'F8', 'T7', 'T8', 'P7', 'P8', 'Fz', 'Pz', 'Oz']
info = mne.create_info(channels, sfreq=100, ch_types='eeg')
info.set_montage('standard_1020')

# Top 10 features
features = ["SP_CHIRP_all_P7", "SP_COUPL_MAG_slow_P7", "SP_DUR_slow_Fp2",
            "SO_TRANS_F3", "SO_NEG_DUR_F3", "SO_POS_DUR_T7", "SO_DUR_F3",
            "SO_TRANS_C3", "SP_COUPL_MAG_all_P8", "SP_ISA_M_slow_F4"]

# Plotting
fig, axes = plt.subplots(2, 5, figsize=(20, 10))
axes = axes.flatten()

for ax, feat in zip(axes, features):
    ch = feat.split('_')[-1]
    diff_array = np.zeros(len(channels))
    diff_array[channels.index(ch)] = df.groupby('group')[feat].mean().diff().iloc[-1]

    mne.viz.plot_topomap(diff_array, info, axes=ax, cmap='RdBu_r', show=False)
    ax.set_title(feat)

plt.suptitle('Spatial Distributions (ASD - HC)', fontsize=12)
plt.tight_layout(rect=[0, 0, 1, 1])
plt.show()
```

\newpage

# Environmental Setup

```{r session_info}
sessionInfo()
```
